Type Refinements for the Working Class

### 2015-12-19 paper.tex
Give meaning explanation for typehood in CTT

\subsection{Parametric Refinement}

@@

\begin{remark}

  At this point, it should be noted that this is very similar to the semantical
  explanation of typehood given in~\cite{martin-lof:1979}, except that we have
  generalized it to a multi-sorted setting, and that we have fibred the entire
  apparatus over collections $\Upsilon$ of symbols.

  In fact, the complexity of the above meaning explanation is an artifact of
  the pointwise style in which we have expressed it. Considered internally to
  the presheaf topos $\Sets^\SCtx$, a refinement is merely a section of the
  object of $\Exprs_\tau$-PERs $\mathfrak{per}(\Exprs_\tau)$, defined
  internally as a subobject of $\Omega^{\Exprs_\tau\times\Exprs_\tau}$:
  \[
    \Define{
      \mathfrak{per}(X)
    }{
      \{ \phi : \Omega^{X\times X}
         \mid
         symmetric(\phi)
         \land transitive(\phi)
      \}
    }
  \]

  So, $\Refines{\Upsilon}{\phi}{\tau}$ obtains just when we have
  $\Member{\phi}{\mathfrak{per}(\Exprs_\tau)(\Upsilon)}$. The benefit of
  explaining such judgments in terms of the presheaf topos is that we do not
  need to deal with the complexities of quantifying over renamings, since this
  is implicit in the definition of the exponential of presheaves,
  \begin{align*}
    \ADefine{
      B^A(\Upsilon)
    }{
      \Hom{\Sets^\SCtx}{
        \Yoneda{\Upsilon}\times A
      }{
        B
      }
    }\\
    &\cong
    \OMode{
      \int_{\Upsilon'}
      (\Upsilon\inj\Upsilon')
      \Rightarrow
      {B(\Upsilon')}^{A(\Upsilon')}
    }
  \end{align*}

@@

\section{Multi-Sorted Computational Type Theory}

@@

>>>>

\subsection{Parametric Refinement}
\label{sec:parametric-refinement}

@@

\begin{remark}\label{rem:internal}

  At this point, it should be noted that this is very similar to the semantical
  explanation of typehood given in~\cite{martin-lof:1979}, except that we have
  generalized it to a multi-sorted setting, and that we have fibred the entire
  apparatus over collections $\Upsilon$ of symbols; furthermore, whereas
  Martin-L\"of defines types in terms of evaluation to canonical form, we have
  remained agnostic on this point as far as refinements are concerned. We will
  come back to the notions of \emph{type} and \emph{computation} in
  section~\ref{sec:ctt}.

  In fact, the complexity of the above meaning explanation is an artifact of
  the pointwise style in which we have expressed it. Considered internally to
  the presheaf topos $\Sets^\SCtx$, a refinement is merely a section of the
  object of $\Exprs_\tau$-PERs $\PERs{\Exprs_\tau}$, defined
  internally as a subobject of $\Omega^{\Exprs_\tau\times\Exprs_\tau}$:
  \[
    \Define{
      \PERs{X}
    }{
      \{ \phi : \Omega^{X\times X}
         \mid
         symmetric(\phi)
         \land transitive(\phi)
      \}
    }
  \]

  So, $\Refines{\Upsilon}{\phi}{\tau}$ obtains just when we have
  $\Member{\phi}{\PERs{\Exprs_\tau}(\Upsilon)}$. The benefit of
  explaining such judgments in terms of the presheaf topos is that we do not
  need to deal with the complexities of quantifying over renamings, since this
  is implicit in the definition of the exponential of presheaves,
  \begin{align*}
    \ADefine{
      B^A(\Upsilon)
    }{
      \Hom{\Sets^\SCtx}{
        \Yoneda{\Upsilon}\times A
      }{
        B
      }
    }\\
    &\cong
    \OMode{
      \int_{\Upsilon'}
      (\Upsilon\inj\Upsilon')
      \Rightarrow
      {B(\Upsilon')}^{A(\Upsilon')}
    }
  \end{align*}

@@

\section{Multi-Sorted Computational Type Theory}
\label{sec:ctt}

@@

\subsection{When is a refinement a type?}

As we remarked in section~\ref{sec:parametric-refinement}, the notion of
\emph{refinement} that we described is very much akin to a multi-sorted
generalization of Martin-L\"of's explanation of types
in~\cite{martin-lof:1979}, except that we have not required an operational
semantics to be present prior to the definition of refinements.

Types, on the other hand, are intrinsically tied up with computation.  In order
to make this concrete, let us first define what it takes for a refinement to be
a type, in the sense of Martin-L\"of---that is, a \emph{value type}. In the
simple case, where we have a global refinement $\GloballyRefines{\phi}{\tau}$,
we will say $\GlobalIsValueType{\phi}{\tau}$ when there exists a PER $\psi$ on
\emph{values} that factors uniquely through $\phi$.

In general, we must explain typehood for parametric refinements
$\Refines{\Upsilon}{\phi}{\tau}$ as well, however; recall from
Remark~\ref{rem:internal} that the definition of $\phi$ consists in a natural
transformation
$\Of{\phi}{\Hom{\Sets^\SCtx}{\Yoneda{\Upsilon}}{\PERs{\Exprs_\tau}}}$.
Therefore, we say that $\IsValueType{\Upsilon}{\phi}{\tau}$ obtains when there
is a unique $\psi$ such that the following diagram commutes, where
$\lambda(\chi)$ is the exponential transpose of $\chi$:
\[
  \IsValueType{\Upsilon}{\phi}{\tau}
  \triangleq
  \begin{tikzcd}[sep=huge]
    \IMode{\Yoneda{\Upsilon}\times\Values_\tau^2}
      \arrow[r, dotted, "\OMode{\lambda(\psi)}"]
      \arrow[d, swap, hook, "\Pair{\ArrId{}}{\IMode{\iota}}"] &
    \IMode{\Omega}\\
    \IMode{\Yoneda{\Upsilon}\times\Exprs_\tau^2}
      \arrow[swap, ur, "\IMode{\lambda(\phi)}"]
  \end{tikzcd}
\]

Contrary to Martin-L\"of, however, we wish to consider a full range of
\emph{computational types}, including those which are not necessarily defined
by their values; a general apparatus for computation-respecting refinements will
allow us to define, for instance, a type for partial functions.

Then, for any refinement $\Refines{\Upsilon}{\phi}{\tau}$, we say
$\IsType{\Upsilon}{\phi}{\tau}$ just when for any
\begin{tikzcd}[cramped]
  \IMode{\Upsilon}\arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
\end{tikzcd}%
, and for all $\Member{L,M,N}{\Exprs_\tau(\Upsilon')}$, from
$\SBinRel{\sim_\tau}{\Upsilon'}{L}{M}$ and $\RMemEq{\Upsilon'}{M}{N}{\phi}$ we
can conclude $\RMemEq{\Upsilon'}{L}{N}{\phi}$.
Stated internally in the topos, the meaning of parametric typehood for
$\phi$ is perhaps a bit more clear:
\[
  \Define{
    \IsType{-}{\phi}{\tau}
  }{
    \prod_{L,M,N\in\Exprs_\tau}
    {\phi(L,N)}^{L\sim_\tau M \times \phi(M,N)}
  }
\]
